### 2022-11-30 paper2.tex
added paper 4
ie and eg

To introduce some notation, let $z_o = \hat \theta_o/\sigma_o$ and
$z_r = \hat \theta_r/\sigma_r$ denote the $z$-statistic of the original and
replication study, respectively. Here $\hat \theta_o$ and $\hat \theta_r$ are
the corresponding effect estimates (assumed to be normally distributed) of the
unknown effect $\theta$ with standard errors $\sigma_o$ and $\sigma_r$,
respectively. The corresponding one-sided $p$-values are denoted by
$p_o=1-\Phi(z_o)$ and $p_r=1-\Phi(z_r)$, respectively, where $\Phi(\cdot)$
denotes the standard normal cumulative distribution function. Let
$c = \sigma_o^2/\sigma_r^2$ denote the variance ratio of the squared standard
errors of the original and replication effect estimates. The squared standard
errors are usually inversely proportional to the sample size of each study, \ie
$\sigma_o^2 = \kappa^2/n_o$ and $\sigma_r^2 = \kappa^2/n_r$ for some unit
variance $\kappa^2$. The variance ratio $c$ can then be identified as the
relative sample size $c=n_r/n_o$. The relative effect size
\begin{equation}\label{eq2:d}
  d = \frac{\hat \theta_r}{\hat \theta_o} = \frac{1}{\sqrt{c}} \frac{z_r}{z_o}
\end{equation}
quantifies the size of the replication effect estimate $\hat \theta_r$ relative
to the original effect estimate $\hat \theta_o$. The corresponding shrinkage of
the replication effect estimate will be denoted as $s=1-d$.

Suppose the original study achieved statistical significance at one-sided level
$\alpha$, so $p_o \leq \alpha$. The standard approach to assess replication
success is based on significance of the replication effect estimate at the same
level $\alpha$, \ie the replication is considered successful if also
$p_r \leq \alpha$. This approach is known in drug development as the two-trials
rule \citep{Senn2008}, usually conducted at $\alpha=0.025$. Let
$z_\alpha = \Phi^{-1}(1-{\alpha})>0$ denote the $z$-value corresponding to the
level $\alpha$, then significance of the replication study is achieved if
$z_r \geq z_\alpha$, which is equivalent to the condition
\begin{equation}\label{eq2:dSig}
   d \geq \frac{z_\alpha}{z_o  \, \sqrt{c}}
\end{equation}
on the relative effect size \eqref{eq2:d}. The right hand-side goes to zero for
increasing $c$, so if the relative sample size $c$ is large enough, significance
of the replication study can be achieved with any arbitrarily small (but
positive) relative effect size $d$. However, declaring replication success when
there is substantial shrinkage is contrary to common sense, as the replication
effect estimate may not reflect an effect size of the same practical relevance
as the original one, despite its statistical significance.

@@

Equation \eqref{eq2:eq.z} can be further transformed to a
condition on the relative effect size \eqref{eq2:d}:
\begin{equation}\label{eq2:res}
  d \geq  \frac{\sqrt{1+c/(K-1)}}{ \sqrt{c K}} \eqqcolon \dmin .
\end{equation}
To achieve replication success, the relative effect size must be at least as
large as the right hand-side of \eqref{eq2:res}, the minimum relative effect size
$\dmin$, a function of $K$ and the relative sample size $c$. If the relative
sample size becomes very large, \ie $c \rightarrow \infty$, we have
$\dmin \downarrow \dinfty$ where
\begin{equation}\label{eq2:smallestpossible}
  \dinfty = 1/\sqrt{K (K-1)}
\end{equation}
is the \textit{limiting relative effect size}. This shows that the minimum
relative effect size $\dmin$ in \eqref{eq2:res} does not go to zero for
increasing $c$, so replication success cannot be achieved if the relative effect
size $d$ is smaller or equal to $\dinfty$, no matter how large the replication
study is. In contrast, the corresponding criterion \eqref{eq2:dSig} of the
two-trials rule can be achieved for any positive relative effect size,
regardless of how small, provided the replication sample size is sufficiently
large.

@@

The condition $p_S \leq \alpha_S$ for replication success at the golden level is
equivalent to $z_S \geq z_\alpha / \sqrt{\varphi}$, \ie
$z_S \sqrt{\varphi} \geq z_\alpha$. In practice it may be preferable to
recalibrate the sceptical $p$-value $p_{S} = 1 - \Phi(z_S)$ to
$ \tilde{p}_S = 1 - \Phi(z_S \sqrt{\varphi})$, which then needs to be compared
to $\alpha$ (rather than $\alpha_S$) to assess replication success and can thus
be interpreted on the same scale as an ordinary $p$-value. For example, the
recalibrated sceptical $p$-value for the replication of \citet{Pyc2010} turns
out to be $\tilde{p}_S=0.061$ and does not lead to replication success at any
level $\alpha<0.061$, including the standard $0.025$ level.

@@

The main difference between the replication success and the two-trials
rule approach concerns how shrinkage of the replication effect
estimate is handled. Figure~\ref{fig2:fig2} illustrates that
shrinkage is penalized in the assessment of replication success, \ie
the original $p$-value needs to be quite small to achieve replication
success for a relative effect size $d<1$. In contrast, significance of
the replication study can be achieved even if there is substantial
shrinkage, provided the replication sample size is large enough.

@@

The approach described so far takes the original study at face-value since it
assumes that $\hat \theta_o$ is equal to the unknown effect size $\theta$. In
practice, however, there are often good reasons to believe that original effect
estimates have a tendency to be inflated (\eg due to publication bias). One way
to address this issue is to base power calculations on a shrunken version of the
original effect estimate, where the amount of shrinkage is guided by domain
knowledge and a risk of bias assessment of the original study. For illustration,
Figure~\ref{fig2:fig3} also shows conditional power based on 20\% shrinkage of
the original effect estimate which reduces the conditional power for all
methods, especially for a relative sample size $c = 1$. Conditional power for
replication success at the golden level can now be pushed towards 100\% only for
$p_o < 0.018$, which can be derived by solving \eqref{eq2:alphaPrime} for
$\alpha$ with $\alpha'=0.025$ and $\dinfty=0.8$. To be able to push conditional
power based on 20\% shrinkage towards 100\% for all $p_o < 0.025$, equation
\eqref{eq2:alphaPrime} would have to be used directly to relax the level from
$\alpha=0.025$ to $\alpha'=0.033$.

@@

\section{Application}\label{sec2:application}
In this section, we illustrate the proposed methodology using data from four
replication projects. All four projects reported effect estimates that were
transformed to correlation coefficients ($r$). This scale allows for easy
comparison of effect estimates from studies that investigate different phenomena
and is bounded to the interval between minus one and one. Moreover, the Fisher
$z$-transformation $\hat{\theta} = \text{tanh}^{-1}(r)$ can be applied to the
correlation coefficients, resulting in the transformed estimates being
asymptotically normal with variance which is only a function of the study sample
size $n$, \ie $\Var(\hat{\theta}) = 1/(n - 3)$ \citep{Fisher1921}.

@@

Figure~\ref{fig2:fig6} displays the relative effect size $d$ versus the original
$p$-value $p_o$ for each study pair and stratified by project. Note that one
study pair from the Philosophy project is not shown due to extremely small
original $p$-value and another study pair from the Psychology project is not
shown due to a very large relative effect size. We can see that for most of the
study pairs, the replication success approach and the two-trials rule lead to
the same conclusion, only six replications show conflicting results. They are
highlighted with diamonds in Figure~\ref{fig2:fig6} and their characteristics are
summarised in Table~\ref{tbl:discrep}.
\begin{table}[!ht]
    \centering
  \caption{Characteristics of studies for which the replication success approach
    (at the golden level) and the two-trials rule disagree (at one-sided
    $\alpha = 0.025$). Shown are relative sample size $c$, relative effect size
    $d$, original, replication and recalibrated sceptical $p$-value $p_o$, $p_r$
    and $\tilde{p}_S$.}
  \label{tbl:discrep}
% \resizebox{\textwidth}{!} {
% latex table generated in R 4.0.5 by xtable 1.8-4 package
% Fri Nov 25 12:12:30 2022
\begin{tabular}{lllllll}
  \toprule
Study & Project & $c$ & $d$ & $p_o$ & $p_r$ & $\tilde{p}_S$ \\
  \midrule
\citet{Schmidt2008} & Psychology & 2.58 & 1.28 & \textbf{0.028} & \textcolor{black}{< 0.0001} & \textcolor{black}{0.024} \\
  \citet{Oberauer2008} & Psychology & 0.60 & 0.67 & \textcolor{black}{0.0003} & \textbf{0.035} & \textcolor{black}{0.017} \\
  \citet{Payne2008} & Psychology & 2.65 & 0.41 & \textcolor{black}{0.001} & \textcolor{black}{0.023} & \textbf{0.031} \\
  \citet{Balafoutas2012} & Social Sciences & 3.48 & 0.52 & \textcolor{black}{0.009} & \textcolor{black}{0.011} & \textbf{0.04} \\
  \citet{Pyc2010} & Social Sciences & 9.18 & 0.38 & \textcolor{black}{0.011} & \textcolor{black}{0.004} & \textbf{0.061} \\
  \citet{Nichols2006} & Experimental Philosophy & 9.40 & 0.49 & \textcolor{black}{0.015} & \textcolor{black}{0.0006} & \textbf{0.049} \\
   \bottomrule
\end{tabular}

% }
\end{table}
Two studies from the Psychology project show replication success but fail the
two-trials rule. These studies show $p$-values that are slightly above the
significance threshold in either original or replication study, but do not
exhibit much shrinkage; in the replication of \citet{Oberauer2008}, the
replication $p$-value was $p_r = 0.035$, a little too large to pass the
two-trials rule. However, as the replication effect estimate shrunk only about
$30$\% compared to the original one, replication success is still achieved.
Conversely, the original $p$-value $p_o = 0.028$ in \citet{Schmidt2008} was just
above the significance level, yet the replication led to a highly significant
result $p_r < 0.0001$ with the effect estimate being even $30$\% larger than the
original counterpart, which therefore also resulted in replication success.

@@

With this paper we further advanced the reverse-Bayes methodology for the
analysis and design of replication studies, yet certain limitations and
opportunities for future research remain: First, assuming normality of the
effect estimates may be questionable, especially for small sample sizes, and
more robust distributional assumptions could be considered. Second, in some
types of analyses (\eg regression or ANOVA) the effect estimate is a vector and
the approach would need suitable adaptations. Third, there is a recent trend to
not only conduct one but several replications for one original study
\citep[\eg][]{Klein2014, Ebersole2016, Klein2018}. Also for this situation, the
method would need to be adapted, \eg the replication estimates could be first
synthesized and an analysis of replication success could be performed
subsequently.

>>>>

To introduce some notation, let $z_o = \hat \theta_o/\sigma_o$ and
$z_r = \hat \theta_r/\sigma_r$ denote the $z$-statistic of the original and
replication study, respectively. Here $\hat \theta_o$ and $\hat \theta_r$ are
the corresponding effect estimates (assumed to be normally distributed) of the
unknown effect $\theta$ with standard errors $\sigma_o$ and $\sigma_r$,
respectively. The corresponding one-sided $p$-values are denoted by
$p_o=1-\Phi(z_o)$ and $p_r=1-\Phi(z_r)$, respectively, where $\Phi(\cdot)$
denotes the standard normal cumulative distribution function. Let
$c = \sigma_o^2/\sigma_r^2$ denote the variance ratio of the squared standard
errors of the original and replication effect estimates. The squared standard
errors are usually inversely proportional to the sample size of each study, \ie{}
$\sigma_o^2 = \kappa^2/n_o$ and $\sigma_r^2 = \kappa^2/n_r$ for some unit
variance $\kappa^2$. The variance ratio $c$ can then be identified as the
relative sample size $c=n_r/n_o$. The relative effect size
\begin{equation}\label{eq2:d}
  d = \frac{\hat \theta_r}{\hat \theta_o} = \frac{1}{\sqrt{c}} \frac{z_r}{z_o}
\end{equation}
quantifies the size of the replication effect estimate $\hat \theta_r$ relative
to the original effect estimate $\hat \theta_o$. The corresponding shrinkage of
the replication effect estimate will be denoted as $s=1-d$.

Suppose the original study achieved statistical significance at one-sided level
$\alpha$, so $p_o \leq \alpha$. The standard approach to assess replication
success is based on significance of the replication effect estimate at the same
level $\alpha$, \ie{} the replication is considered successful if also
$p_r \leq \alpha$. This approach is known in drug development as the two-trials
rule \citep{Senn2008}, usually conducted at $\alpha=0.025$. Let
$z_\alpha = \Phi^{-1}(1-{\alpha})>0$ denote the $z$-value corresponding to the
level $\alpha$, then significance of the replication study is achieved if
$z_r \geq z_\alpha$, which is equivalent to the condition
\begin{equation}\label{eq2:dSig}
   d \geq \frac{z_\alpha}{z_o  \, \sqrt{c}}
\end{equation}
on the relative effect size \eqref{eq2:d}. The right hand-side goes to zero for
increasing $c$, so if the relative sample size $c$ is large enough, significance
of the replication study can be achieved with any arbitrarily small (but
positive) relative effect size $d$. However, declaring replication success when
there is substantial shrinkage is contrary to common sense, as the replication
effect estimate may not reflect an effect size of the same practical relevance
as the original one, despite its statistical significance.

@@

Equation \eqref{eq2:eq.z} can be further transformed to a
condition on the relative effect size \eqref{eq2:d}:
\begin{equation}\label{eq2:res}
  d \geq  \frac{\sqrt{1+c/(K-1)}}{ \sqrt{c K}} \eqqcolon \dmin .
\end{equation}
To achieve replication success, the relative effect size must be at least as
large as the right hand-side of \eqref{eq2:res}, the minimum relative effect size
$\dmin$, a function of $K$ and the relative sample size $c$. If the relative
sample size becomes very large, \ie{} $c \rightarrow \infty$, we have
$\dmin \downarrow \dinfty$ where
\begin{equation}\label{eq2:smallestpossible}
  \dinfty = 1/\sqrt{K (K-1)}
\end{equation}
is the \textit{limiting relative effect size}. This shows that the minimum
relative effect size $\dmin$ in \eqref{eq2:res} does not go to zero for
increasing $c$, so replication success cannot be achieved if the relative effect
size $d$ is smaller or equal to $\dinfty$, no matter how large the replication
study is. In contrast, the corresponding criterion \eqref{eq2:dSig} of the
two-trials rule can be achieved for any positive relative effect size,
regardless of how small, provided the replication sample size is sufficiently
large.

@@

The condition $p_S \leq \alpha_S$ for replication success at the golden level is
equivalent to $z_S \geq z_\alpha / \sqrt{\varphi}$, \ie{}
$z_S \sqrt{\varphi} \geq z_\alpha$. In practice it may be preferable to
recalibrate the sceptical $p$-value $p_{S} = 1 - \Phi(z_S)$ to
$ \tilde{p}_S = 1 - \Phi(z_S \sqrt{\varphi})$, which then needs to be compared
to $\alpha$ (rather than $\alpha_S$) to assess replication success and can thus
be interpreted on the same scale as an ordinary $p$-value. For example, the
recalibrated sceptical $p$-value for the replication of \citet{Pyc2010} turns
out to be $\tilde{p}_S=0.061$ and does not lead to replication success at any
level $\alpha<0.061$, including the standard $0.025$ level.

@@

The main difference between the replication success and the two-trials
rule approach concerns how shrinkage of the replication effect
estimate is handled. Figure~\ref{fig2:fig2} illustrates that
shrinkage is penalized in the assessment of replication success, \ie{}
the original $p$-value needs to be quite small to achieve replication
success for a relative effect size $d<1$. In contrast, significance of
the replication study can be achieved even if there is substantial
shrinkage, provided the replication sample size is large enough.

@@

The approach described so far takes the original study at face-value since it
assumes that $\hat \theta_o$ is equal to the unknown effect size $\theta$. In
practice, however, there are often good reasons to believe that original effect
estimates have a tendency to be inflated (\eg{} due to publication bias). One way
to address this issue is to base power calculations on a shrunken version of the
original effect estimate, where the amount of shrinkage is guided by domain
knowledge and a risk of bias assessment of the original study. For illustration,
Figure~\ref{fig2:fig3} also shows conditional power based on 20\% shrinkage of
the original effect estimate which reduces the conditional power for all
methods, especially for a relative sample size $c = 1$. Conditional power for
replication success at the golden level can now be pushed towards 100\% only for
$p_o < 0.018$, which can be derived by solving \eqref{eq2:alphaPrime} for
$\alpha$ with $\alpha'=0.025$ and $\dinfty=0.8$. To be able to push conditional
power based on 20\% shrinkage towards 100\% for all $p_o < 0.025$, equation
\eqref{eq2:alphaPrime} would have to be used directly to relax the level from
$\alpha=0.025$ to $\alpha'=0.033$.

@@

\section{Application}\label{sec2:application}
In this section, we illustrate the proposed methodology using data from four
replication projects. All four projects reported effect estimates that were
transformed to correlation coefficients ($r$). This scale allows for easy
comparison of effect estimates from studies that investigate different phenomena
and is bounded to the interval between minus one and one. Moreover, the Fisher
$z$-transformation $\hat{\theta} = \text{tanh}^{-1}(r)$ can be applied to the
correlation coefficients, resulting in the transformed estimates being
asymptotically normal with variance which is only a function of the study sample
size $n$, \ie{} $\Var(\hat{\theta}) = 1/(n - 3)$ \citep{Fisher1921}.

@@

Figure~\ref{fig2:fig6} displays the relative effect size $d$ versus the original
$p$-value $p_o$ for each study pair and stratified by project. Note that one
study pair from the Philosophy project is not shown due to extremely small
original $p$-value and another study pair from the Psychology project is not
shown due to a very large relative effect size. We can see that for most of the
study pairs, the replication success approach and the two-trials rule lead to
the same conclusion, only six replications show conflicting results. They are
highlighted with diamonds in Figure~\ref{fig2:fig6} and their characteristics are
summarised in Table~\ref{tbl:discrep}.
\begin{table}[!ht]
    \centering
  \caption{Characteristics of studies for which the replication success approach
    (at the golden level) and the two-trials rule disagree (at one-sided
    $\alpha = 0.025$). Shown are relative sample size $c$, relative effect size
    $d$, original, replication and recalibrated sceptical $p$-value $p_o$, $p_r$
    and $\tilde{p}_S$.}
  \label{tbl:discrep}
\resizebox{\textwidth}{!} {
% latex table generated in R 4.0.5 by xtable 1.8-4 package
% Fri Nov 25 12:12:30 2022
\begin{tabular}{lllllll}
  \toprule
Study & Project & $c$ & $d$ & $p_o$ & $p_r$ & $\tilde{p}_S$ \\
  \midrule
\citet{Schmidt2008} & Psychology & 2.58 & 1.28 & \textbf{0.028} & \textcolor{black}{< 0.0001} & \textcolor{black}{0.024} \\
  \citet{Oberauer2008} & Psychology & 0.60 & 0.67 & \textcolor{black}{0.0003} & \textbf{0.035} & \textcolor{black}{0.017} \\
  \citet{Payne2008} & Psychology & 2.65 & 0.41 & \textcolor{black}{0.001} & \textcolor{black}{0.023} & \textbf{0.031} \\
  \citet{Balafoutas2012} & Social Sciences & 3.48 & 0.52 & \textcolor{black}{0.009} & \textcolor{black}{0.011} & \textbf{0.04} \\
  \citet{Pyc2010} & Social Sciences & 9.18 & 0.38 & \textcolor{black}{0.011} & \textcolor{black}{0.004} & \textbf{0.061} \\
  \citet{Nichols2006} & Experimental Philosophy & 9.40 & 0.49 & \textcolor{black}{0.015} & \textcolor{black}{0.0006} & \textbf{0.049} \\
   \bottomrule
\end{tabular}%
}
\end{table}
Two studies from the Psychology project show replication success but fail the
two-trials rule. These studies show $p$-values that are slightly above the
significance threshold in either original or replication study, but do not
exhibit much shrinkage; in the replication of \citet{Oberauer2008}, the
replication $p$-value was $p_r = 0.035$, a little too large to pass the
two-trials rule. However, as the replication effect estimate shrunk only about
$30$\% compared to the original one, replication success is still achieved.
Conversely, the original $p$-value $p_o = 0.028$ in \citet{Schmidt2008} was just
above the significance level, yet the replication led to a highly significant
result $p_r < 0.0001$ with the effect estimate being even $30$\% larger than the
original counterpart, which therefore also resulted in replication success.

@@

With this paper we further advanced the reverse-Bayes methodology for the
analysis and design of replication studies, yet certain limitations and
opportunities for future research remain: First, assuming normality of the
effect estimates may be questionable, especially for small sample sizes, and
more robust distributional assumptions could be considered. Second, in some
types of analyses (\eg{} regression or ANOVA) the effect estimate is a vector and
the approach would need suitable adaptations. Third, there is a recent trend to
not only conduct one but several replications for one original study
\citep[\eg{}][]{Klein2014, Ebersole2016, Klein2018}. Also for this situation, the
method would need to be adapted, \eg{} the replication estimates could be first
synthesized and an analysis of replication success could be performed
subsequently.
